### 2017-06-16 Paper Drafts/Final/finalDraft.tex
Final Draft

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Review Volume (last updated on 20-4-2015)                            %%
%% Trim Size: 9in x 6in                                                 %%
%% Text Area: 7.35in (include runningheads) x 4.5in                     %%
%% Main Text: 10 on 13pt                                                %%
%% For support: Yolande Koh, <ykoh@wspc.com.sg>                         %%
%%              D. Rajesh Babu, <rajesh@wspc.com.sg>                    %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
\documentclass{ws-rv9x6}
\usepackage{subfigure}   % required only when side-by-side / subfigures are used
\usepackage{ws-rv-thm}   % comment this line when `amsthm / theorem / ntheorem` package is used
\usepackage{ws-rv-van}   % numbered citation & references (default)
%\usepackage{ws-index}   % to produce multiple indexes
\makeindex
%\newindex{aindx}{adx}{and}{Author Index}       %author index
%\renewindex{default}{idx}{ind}{Subject Index}  %subject index

@@

\author[J.P. Meagher et al.]{J.P. Meagher\footnote{Department of Statistics, The University of Warwick}, T. Damoulas\footnote{The University of Warwick and The Alan Turing Institute}, K.E. Jones\footnote{Centre for Biodiversity and Environment Research, University College London}, and M. Girolami\footnote{Imperial College London and The Alan Turing Institute}\\}
%\index[aindx]{Author, F.} % or \aindx{Author, F.}
%\index[aindx]{Author, S.} % or \aindx{Author, S.}

\address{
J.Meagher@Warwick.ac.uk}

@@

%\tableofcontents
\section{Introduction}
\label{sec:intro}
The emerging field of Data Science is driven by research which lies at the nexus of Statistics and Computer Science. Bioacoustics is one such area generating vast quantities of data, often through citizen science initiatives. \cite{allen2006citizen} \cite{pettorelli2013indicator} Bioacoustic techniques for biodiversity monitoring \cite{stathopoulos2017bat} \cite{damoulas2010bayesian} have the potential to make real policy impacts, particularly with regard to sustainable economic development and nature conservation.

Bats (order \textit{Chiroptera}) have been identified as ideal bioindicators for monitoring climate change and habitat quality,\cite{jones2009carpe} and are of particular interest for monitoring biodiversity acoustically. Typically, a bat broadcasts information about itself in an ultrasonic echolocation call.\cite{griffin1944echolocation} The development of automatic acoustic monitoring algorithms \cite{stathopoulos2017bat} \cite{walters2012continental} means that large scale, non-invasive monitoring of bats is becoming possible.

Monitoring bat populations provides useful information, but an understanding the orders evolutionary history is required to identify the cause and effect of any changes observed. The echolocation call structure, which reflects a bats diet and habitat,\cite{aldridge1987morphology}  is a key aspect of this natural history. Reconstructing ancestral traits\cite{joy2016ancestral} relies on a statistical comparative analysis incorporating extant species and fossil records. \cite{felsenstein2004inferring} However, the fossil record is of limited use in inferring ancestral echolocation calls in bats. Therefore, statistical data science techniques may shed some light on this topic.

Previous studies of bat echolocation calls for both classification \cite{walters2012continental} and ancestral reconstruction \cite{collen2012evolution} analysed features extracted from the call spectrogram. These call features relied upon domain knowledge to ensure they were sensibly selected and applied. More recently, general techniques for the classification of acoustic signals have been developed. \cite{stathopoulos2014bat} \cite{damoulas2010bayesian} General techniques for the ancestral reconstruction of function-valued traits have also been proposed. \cite{group2012phylogenetic} Jones \& Moriarty \cite{jones2013evolutionary} extend Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits\cite{meyer2005up} over a phylogeny, a method which was demonstrated for synthetic data by Hajipantelis et al.\cite{hadjipantelis2013function}. This current research investigates these techniques in the context of the bat echolocation calls.

Section \ref{sec:funrep} presents details on representing echolocation call recordings as function-valued traits, and Section \ref{sec:pgp} then illustrates the model for evolution developed by a phylogenetic Gaussian Process. The results of our analysis of bat echolocation calls are then presented and discussed. 
\section{Echolocation Calls as Function-Valued Traits}
\label{sec:funrep}
A functional data object is generated when repeated measurements of some process are taken along a continuous scale, such as time. \cite{ramsay2006functional} 
These measurements can be thought of as representing points on a curve that varies gradually and continuously. In the context of phylogenetics, these functional data objects are function-valued traits. \cite{meyer2005up}

@@

Given a phylogenetic tree \(\mathbf{T}\), representing the evolutionary relationships between the recorded bat species, we denote the \(m^{th}\) call recording of the \(l^{th}\) individual bat of the species observed at point \(\mathbf{t} \in \mathbf{T}\) by \(\{\hat{x}_{lm}^{\mathbf{t}}(n)\}_{n=0}^{N_{lm}^{\mathbf{t}} - 1}\). 
Thus, \(\{\hat{x}_{lm}^{\mathbf{t}}(n)\}\) is a series of discrete measurements of the function \(x_{lm}^{\mathbf{t}}(\cdot)\), observed at the time points given by \(\frac{n}{f_S}\), where \(f_S\) is the sampling rate, in samples per second (Hz), of the recording. Assume then that \(x_{lm}^{\mathbf{t}}(\cdot) = x_{l}^{\mathbf{t}}(\cdot) + z_{lm}^{\mathbf{t}}(\cdot)\), where \(x_{l}^{\mathbf{t}}(\cdot)\) is the representative call function for the \(l^{th}\) individual and \(z_{lm}^{\mathbf{t}}(\cdot)\) is the noise process for the \(m^{th}\) call. Further, assume that \(x_{l}^{\mathbf{t}}(\cdot) = x^{\mathbf{t}}(\cdot) + z_{l}^{\mathbf{t}}(\cdot)\) where \(x^{\mathbf{t}}(\cdot)\) is the representative call function for the bat species at \({\mathbf{t}}\) and \(z_{l}^{\mathbf{t}}(\cdot)\) is the noise process for the \(l^{th}\) individual. It is the echolocation call functions at the species level that we re interested in.

The call recordings themselves are functional data objects, however modelling the phylogenetic relationships between \(\{x_{lm}^{\mathbf{t}}(t)\}\) and \(\{x_{l'm'}^{{\mathbf{t}}'}(t)\}\) directly implies that the processes are comparable at time \(t\). 
This is not the case for acoustic signals, a phenomenon which is often addressed by dynamic time warping.\cite{berndt1994using} Another approach to this issue is to consider an alternative functional representation of the signal.

The Fourier transform of \(x_{lm}^{\mathbf{t}}(\cdot)\) is given by
\[
X_{lm}^{\mathbf{t}}(f) = \int_{-\infty}^{\infty} x_{lm}^{\mathbf{t}}(t) e^{-i 2\pi f t} dt.
\label{eqn:dft}
\]
The energy spectral density of \(x_{lm}^{\mathbf{t}}(\cdot)\) is the magnitude of the Fourier transform and the log energy spectral density is given by
\[
\mathcal{E}_{lm}^{\mathbf{t}}(\cdot) = 10 \log_{10} \left( |X_{lm}^{\mathbf{t}}(\cdot)|^2 \right).
\]

@@

The discrete Fourier Transform\cite{antoniou2006digital}
of \(\{\hat{x}_{lm}^{\mathbf{t}}(n)\}\) provides an estimate for the log energy spectral density, the positive frequencies of which are denoted \(\{\mathcal{E}_{lm}^{\mathbf{t}}(k) : k = 0, \dots, \frac{ N_{lm}^{\mathbf{t}}}{2} + 1\}\). Smoothing splines\cite{friedman2001elements} are applied to this series to obtain \(\hat{\mathcal{E}}_{lm}^{\mathbf{t}}(\cdot)\), a smooth function estimating \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\).

@@

\section{Phylogenetic Gaussian Processes}
\label{sec:pgp}
A Gaussian process places a prior distribution over functions, \(f(x) \sim \mathcal{GP}(m(x), k (x, x'),\) where \(x \in \mathbf{R}^P\) is some input variable, the mean function \(m(x) = \mathbf{E}[f(x)]\), and the covariance function \(k(x, x') = \mathnormal{cov}(f(x), f(x') )\). Given observations \(\mathbf{y}\) at locations \(\{x_n\}_{n=1}^N\), Gaussian noise, and kernel hyperparameters \(\theta\), a posterior predictive distribution over functions can be inferred analytically. See Rasmussen \& Williams\cite{rasmussen2006gaussian} for an in depth treatment.

Jones \& Moriarty\cite{jones2013evolutionary} extend GPs for the inference of function-valued traits over a phylogeny. Consider \(\mathcal{E}^{\mathbf{t}}(\cdot)\), a functional representation of the echolocation call of the species observed at the point \(\mathbf{t}\) on the phylogenetic tree \(\mathbf{T}\) with respect to frequency. Modelling this as GP function, where \(\mathcal{E}^{\mathbf{t}}(f)\) corresponds to a point \((f, \mathbf{t})\) on the frequency-phylogeny \(F \times \mathbf{T}\), requires that a suitable phylogenetic covariance function, \(\Sigma_{\mathbf{T}}\left((f,\mathbf{t}), (f',\mathbf{t}')\right)\), is defined.

@@

The problem of obtaining estimators \(\hat{\phi}_F\) and \(\hat{X}_{\mathbf{T}}\) is dealt with by Hajipantelis et al. \cite{hadjipantelis2013function} \(\hat{\phi}_F\) is obtained by Independent Components Analysis, as described by Blaschke \& Wiscott\cite{blaschke2004cubica} after using a resampling procedure to obtain stable principal components for the observed traits. Given \(\hat{\phi}_F\), the estimated matrix of mixing coefficients is \(\hat{X}_{\mathbf{T}} = \mathcal{E} (\hat{\phi}_F^{\mathsf{T}})^{-1}\).

@@

We note that the elements of \(\theta\) each have intuitive interpretations. The variance of observations in the sample is \(\sigma_p + \sigma_n\), where \(\sigma_p\) is the phylogenetic noise, and \(\sigma_n\) is the non-phylogenetic noise. \(\sigma_p\) is the proportion of variation which can be accounted for by the distance between any \(\mathbf{t}, \mathbf{t}' \in \mathbf{T}\), while \(\sigma_n\) accounts for other sources of variation which may be observed at \(\mathbf{t}\). The length-scale parameter, \(\ell\) then indicates the strength of the correlation along \(\mathbf{T}\), where large values of \(\ell\) indicate a correlation that decays slowly as \(d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')\) increases.

\section{Results}

\subsection{Data Description}

The post processed echolocation call data for bats found across north and central Mexico which accompanied Stathopoulos et al. \cite{stathopoulos2017bat} was used in this analysis. Live bats were caught, identified, and recorded either while released from the hand or while tied to a zip line. Calls were recorded at a sampling frequency of 500 kHz. 
In total the dataset consists of 22 species in five families, 449 individual bats and 1816 individual echolocation call recordings. The distribution of these call recordings across species is summarised in Table \ref{tab::dataset}.

Collen's \cite{collen2012evolution} Bat super-tree provided the basis for the phylogenetic tree of the recorded bat species, \(\mathbf{T}\).

@@

We are interested in modelling the evolution of \(\mathcal{E}^{\mathbf{t}}(\cdot)\), the function valued trait representing \(x_{lm}^{\mathbf{t}}(\cdot)\). However, only 22 species of bat are represented in \(\mathbf{T}\). The relatively small size of this dataset presents challenges for the estimation of the kernel hyperparameters in (\ref{eqn:oukernel}). A short simulation study was performed to investigate the dynamics at play.

We are not limited to a single observation at any given \(\mathbf{t}\). By repeatedly sampling at each observed \(\mathbf{t}\), larger samples can be obtained, improving the quality of the estimators \(\hat{\theta}\). With this in mind, 1000 independent, univariate phylogenetic GPs were simulated for each of \(n = \{1,2,4,8\}\) according to the kernel (\ref{eqn:oukernel}) with \(\theta = [1,50,1]^{\mathsf{T}}\), where \(n\) is the number of samples generated at each leaf node. The likelihood of each of these samples (\ref{eqn:t2mle}) is then maximised to give a type II maximum likelihood estimator \(\hat{\theta}\) and the results summarised in Table \ref{tab::ind_simulation}. This simulation study indicates that at least \(n = 4\) observations are needed at each leaf node to provide stable estimators \(\hat{\theta}\). Thus, when implementing the phylogenetic OU process for bat echolocation call traits, resampling methods must be used to provide multiple estimates for \(\mathcal{E}^{\mathbf{t}}(\cdot)\)

@@

Given the modelling assumptions made in Section \ref{sec:funrep} the best estimator for \(\mathcal{E}^{\mathbf{t}}(\cdot)\) is the sample mean given by 
\begin{equation}
\hat{\mathcal{E}}^{\mathbf{t}}~(\cdot) = \frac{1}{l_{\mathbf{t}}} \sum_{l = 1}^{l_{\mathbf{t}}} \frac{1}{m_l} \sum_{m = 1}^{m_l} \hat{\mathcal{E}}_{lm}^{\mathbf{t}}(\cdot)
\label{eqn:est}
\end{equation}
where \(m_l\) is the total number of recordings for the \(l^{th}\) individual and \(l_{\mathbf{t}}\) is the number of individuals recorded from the species at \(\mathbf{t} \in \mathbf{T}\). However, Table \ref{tab::ind_simulation} indicates that 22 samples is not enough to obtain a stable \(\hat{\theta}\) by type II maximum likelihood estimation. We implement a resampling procedure to leverage more estimates from the dataset in order to produce more stable \(\hat{\theta}\).

@@

Our analysis took \(n_r = 4\) and included 4 samples of \(\hat{\mathcal{E}}_r^{\mathbf{t}}(\mathbf{f})\) in each \(\hat{\mathcal{E}}_r\). This reflected the structure of the dataset, for which the minimum number of individuals per species was 4, and the results of the simulations study which showed that 4 observations per species provided reasonably stable estimates for \(\theta\). Note also that \(\mathbf{f} = [9,10, \dots, 212]^{\mathsf{T}}\), which reflects the spectrum of frequencies over which bats emit echolocation calls. \(\hat{\phi}_F\) was obtained by averaging the basis identified over all \(\hat{\mathcal{E}}_r\) and applying a single set of basis functions to each dataset. Thus \(\hat{X}_r\), the matrix of mixing coefficients described by (\ref{eqn:inv}), the columns of which are modelled a phylogenetic OU processes, is obtained for each \(\hat{\mathcal{E}}_r\). \(\hat{\theta}_r\) is then the type II maximum likelihood estimator of (\ref{eqn:t2mle}) given \(\hat{\mathcal{E}}_r\). Table \ref{tab::bat_results} presents the results of the hyperparameter estimation procedure.

@@

To perform ancestral trait reconstruction for \(\mathcal{E}^{\mathbf{t}^*}(\cdot)\) the species level traits are estimated by (\ref{eqn:est}) and and the model hyperparameters by the mean values of \(\theta_r\) reported in Table \ref{tab::bat_results}.

@@

\section{Conclusions and Further Work}

This preliminary analysis has developed a limited model for the evolution of echolocation in bats and identified a phylogenetic signal which allows the construction of a posterior predictive distribution for ancestral traits. The log energy spectral density has been identified as a trait representative of the echolocation call in bats. This trait, representing the energy intensity of the call across the frequency spectrum, is modelled as a series of independent components, combinations of energy intensities across the spectrum, each of which evolves according to a phylogenetic Ornstein-Uhlenbeck process. Estimating the hyperparameters governing these Ornstein-Uhlenbeck processes from observed traits provides an insight into the evolutionary path of these traits. Each of the hyperparameters has an intuitive interpretation where \(\frac{\sigma_p}{\sigma_p + \sigma_n}\) indicates the proportion of variation in the sample attributable solely to the evolutionary path while \(\ell\) provides a measure of how quickly correlation between observations along the evolutionary path decays. We are working towards understanding what the results of this analysis could mean with respect to the evolution of echolocation in bats.

One particular limitation of the model is the representation of the echolocation call by a log energy spectral density. Echolocation calls have complex spectral and temporal structures, much of which is lost in the log energy spectral density representation. An alternative time-frequency representation, which preserves more of this structure, is the spectrogram. Modelling the evolution of bat echolocation calls with spectrograms is an active area of future research.

\bibliographystyle{ws-rv-van}
\bibliography{../BatBiblio}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Review Volume (last updated on 20-4-2015)                            %%
%% Trim Size: 9in x 6in                                                 %%
%% Text Area: 7.35in (include runningheads) x 4.5in                     %%
%% Main Text: 10 on 13pt                                                %%
%% For support: Yolande Koh, <ykoh@wspc.com.sg>                         %%
%%              D. Rajesh Babu, <rajesh@wspc.com.sg>                    %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%\documentclass[wsdraft]{ws-rv9x6} % to draw border line around text area
\documentclass{ws-rv9x6}
\usepackage{subfigure}   % required only when side-by-side / subfigures are used
\usepackage{ws-rv-thm}   % comment this line when `amsthm / theorem / ntheorem` package is used
\usepackage[square]{ws-rv-van}   % numbered citation & references (default)
%\usepackage{ws-index}   % to produce multiple indexes
\makeindex
%\newindex{aindx}{adx}{and}{Author Index}       %author index
%\renewindex{default}{idx}{ind}{Subject Index}  %subject index

@@

\author[J.P. Meagher et al.]{J.P. Meagher\footnote{J.P Meagher would like to thank the EPSRC for funding this work, and also acknowledge the support of the Alan Turing Institute and Gaelic Players Association.}, T. Damoulas\footnote{The University of Warwick and The Alan Turing Institute}, K.E. Jones\footnote{Centre for Biodiversity and Environment Research, University College London}, and M. Girolami\footnote{Imperial College London and The Alan Turing Institute}\\}
%\index[aindx]{Author, F.} % or \aindx{Author, F.}
%\index[aindx]{Author, S.} % or \aindx{Author, S.}

\address{Department of Statistics,\\ The University of Warwick.\\
	J.Meagher@Warwick.ac.uk}

@@

%\tableofcontents
\section{Introduction}
\label{sec:intro}
The emerging field of Data Science is driven by research which lies at the nexus of Statistics and Computer Science. Bioacoustics is one such area generating vast quantities of data, often through citizen science initiatives \cite{pettorelli2013indicator}. Bioacoustic techniques for biodiversity monitoring \cite{stathopoulos2017bat} \cite{damoulas2010bayesian} have the potential to make real policy impacts, particularly with regard to sustainable economic development and nature conservation.

Bats (order \textit{Chiroptera}) have been identified as ideal bioindicators for monitoring climate change and habitat quality \cite{jones2009carpe}, and are of particular interest for monitoring biodiversity acoustically. Typically, a bat broadcasts information about itself in an ultrasonic echolocation call \cite{griffin1944echolocation}. The development of automatic acoustic monitoring algorithms \cite{stathopoulos2017bat} \cite{walters2012continental} means that large scale, non-invasive monitoring of bats is becoming possible.

Monitoring bat populations provides useful information, but an understanding the orders evolutionary history is required to identify the cause and effect of any changes observed. The echolocation call structure, which reflects a bats diet and habitat \cite{aldridge1987morphology}, is a key aspect of this evolutionary history. Reconstructing ancestral traits \cite{joy2016ancestral} relies on a statistical comparative analysis incorporating extant species and fossil records \cite{felsenstein2004inferring}. However, the fossil record is of limited use in inferring ancestral echolocation calls in bats. Therefore, statistical data science techniques may shed some light on this topic.

Previous studies of bat echolocation calls for both classification \cite{walters2012continental} and ancestral reconstruction \cite{collen2012evolution} analysed features extracted from the call spectrogram. These call features relied upon domain knowledge to ensure they were sensibly selected and applied. More recently, general techniques for the classification of acoustic signals have been developed \cite{stathopoulos2014bat} \cite{damoulas2010bayesian}. General techniques for the ancestral reconstruction of function-valued traits have also been proposed \cite{group2012phylogenetic}. Jones \& Moriarty \cite{jones2013evolutionary} extend Gaussian Process Regression \cite{rasmussen2006gaussian} to model the evolution of function-valued traits \cite{meyer2005up} over a phylogeny, a method which was demonstrated for synthetic data by Hajipantelis et al \cite{hadjipantelis2013function}. This current research investigates these techniques in the context of the bat echolocation calls.

The structure of this paper is as follows, section \ref{sec:funrep} presents details on representing echolocation call recordings as function-valued traits. Section \ref{sec:pgp} develops the model for evolution given by a phylogenetic Gaussian Process. The results of the analysis of bat echolocation calls are then presented and discussed in sections \ref{sec:results} and \ref{sec:conc}.

\section{Echolocation Calls as Function-Valued Traits}
\label{sec:funrep}
A functional data object is generated when repeated measurements of some process are taken along a continuous scale, such as time \cite{ramsay2006functional}. 
These measurements can be thought of as representing points on a curve that varies gradually and continuously. In the context of phylogenetics, these functional data objects are function-valued traits \cite{meyer2005up}.

@@

Given a phylogenetic tree \(\mathbf{T}\), representing the evolutionary relationships between the recorded bat species, we denote the \(m^{th}\) call recording of the \(l^{th}\) individual bat of the species observed at point \(\mathbf{t} \in \mathbf{T}\) by \(\{\hat{x}_{lm}^{\mathbf{t}}(n)\}_{n=0}^{N_{lm}^{\mathbf{t}} - 1}\). 
Thus, \(\{\hat{x}_{lm}^{\mathbf{t}}(n)\}\) is a series of discrete measurements of the function \(x_{lm}^{\mathbf{t}}(\cdot)\), observed at the time points given by \(\frac{n}{f_S}\), where \(f_S\) is the sampling rate, in samples per second (Hz), of the recording. Assume then that \(x_{lm}^{\mathbf{t}}(\cdot) = x_{l}^{\mathbf{t}}(\cdot) + z_{lm}^{\mathbf{t}}(\cdot)\), where \(x_{l}^{\mathbf{t}}(\cdot)\) is the representative call function for the \(l^{th}\) individual and \(z_{lm}^{\mathbf{t}}(\cdot)\) is the noise process for the \(m^{th}\) call. Further, assume that \(x_{l}^{\mathbf{t}}(\cdot) = x^{\mathbf{t}}(\cdot) + z_{l}^{\mathbf{t}}(\cdot)\) where \(x^{\mathbf{t}}(\cdot)\) is the representative call function for the bat species at \({\mathbf{t}}\) and \(z_{l}^{\mathbf{t}}(\cdot)\) is the noise process for the \(l^{th}\) individual. It is the phylogenetic relationship between the species level echolocation call functions that we are interested in modelling.

The call recordings themselves are functional data objects, however modelling the phylogenetic relationships between \(\{x_{lm}^{\mathbf{t}}(t)\}\) and \(\{x_{l'm'}^{{\mathbf{t}}'}(t)\}\) directly implies that the processes are comparable at time \(t\). 
This is not the case for acoustic signals, a phenomenon which is often addressed by dynamic time warping \cite{berndt1994using}. Another approach to this issue is to consider an alternative functional representation of the signal.

The Fourier transform of \(x_{lm}^{\mathbf{t}}(\cdot)\) is given by
\[
X_{lm}^{\mathbf{t}}(f) = \int_{-\infty}^{\infty} x_{lm}^{\mathbf{t}}(t) e^{-i 2\pi f t} dt.
\label{eqn:dft}
\]
The energy spectral density of \(x_{lm}^{\mathbf{t}}(\cdot)\) is the squared magnitude of the Fourier transform and the log energy spectral density is given by
\[
\mathcal{E}_{lm}^{\mathbf{t}}(\cdot) = 10 \log_{10} \left( |X_{lm}^{\mathbf{t}}(\cdot)|^2 \right).
\]

@@

The discrete Fourier Transform \cite{antoniou2006digital}
of \(\{\hat{x}_{lm}^{\mathbf{t}}(n)\}\) provides an estimate for the log energy spectral density, the positive frequencies of which are denoted \(\{\mathcal{E}_{lm}^{\mathbf{t}}(k) : k = 0, \dots, \frac{ N_{lm}^{\mathbf{t}}}{2} + 1\}\). Smoothing splines \cite{friedman2001elements} are applied to this series to obtain \(\hat{\mathcal{E}}_{lm}^{\mathbf{t}}(\cdot)\), a smooth function estimating \(\mathcal{E}_{lm}^{\mathbf{t}}(\cdot)\).

@@

\section{Phylogenetic Gaussian Processes}
\label{sec:pgp}
A Gaussian process places a prior distribution over functions, \(f(x) \sim \mathcal{GP}(m(x), k (x, x'))\), where \(x \in \mathbf{R}^P\) is some input variable, the mean function \(m(x) = \mathbf{E}[f(x)]\), and the covariance function \(k(x, x') = \mathrm{cov}(f(x), f(x') )\). Given observations \(\mathbf{y}\) at locations \(\{x_n\}_{n=1}^N\), Gaussian noise, and kernel hyperparameters \(\theta\), a posterior predictive distribution over functions can be inferred analytically. See Rasmussen \& Williams \cite{rasmussen2006gaussian} for an in depth treatment.

Jones \& Moriarty \cite{jones2013evolutionary} extend GPs for the inference of function-valued traits over a phylogeny. Consider \(\mathcal{E}^{\mathbf{t}}(\cdot)\), a functional representation of the echolocation call of the species observed at the point \(\mathbf{t}\) on the phylogenetic tree \(\mathbf{T}\) with respect to frequency. Modelling this as GP function, where \(\mathcal{E}^{\mathbf{t}}(f)\) corresponds to a point \((f, \mathbf{t})\) on the frequency-phylogeny \(F \times \mathbf{T}\), requires that a suitable phylogenetic covariance function, \(\Sigma_{\mathbf{T}}\left((f,\mathbf{t}), (f',\mathbf{t}')\right)\), is defined.

@@

The problem of obtaining estimators \(\hat{\phi}_F\) and \(\hat{X}_{\mathbf{T}}\) is dealt with by Hajipantelis et al \cite{hadjipantelis2013function}. \(\hat{\phi}_F\) is obtained by Independent Components Analysis, as described by Blaschke \& Wiscott \cite{blaschke2004cubica} after using a resampling procedure to obtain stable principal components for the observed traits. Given \(\hat{\phi}_F\), the estimated matrix of mixing coefficients is \(\hat{X}_{\mathbf{T}} = \mathcal{E} (\hat{\phi}_F^{\mathsf{T}})^{-1}\).

@@

We note that the elements of \(\theta\) each have intuitive interpretations. The total variation of observed points is \(\sigma_p + \sigma_n\), where \(\sigma_p\) is the phylogenetic noise, and \(\sigma_n\) is the non-phylogenetic noise. \(\sigma_p\) is the variation depending on the evolutionary time between any \(\mathbf{t}, \mathbf{t}' \in \mathbf{T}\), while \(\sigma_n\) accounts for variation that does not depend on the phylogeny. The length-scale parameter, \(\ell\), indicates the strength of the correlation between points on \(\mathbf{T}\), where large values of \(\ell\) indicate a correlation that decays slowly as \(d_{\mathbf{T}}(\mathbf{t}, \mathbf{t}')\) increases.

\section{Results}
\label{sec:results}
\subsection{Data Description}

Post processed echolocation call data accompanying Stathopoulos et al. \cite{stathopoulos2017bat} was used in this analysis. Live bats were caught, identified, and recorded in at a sampling frequency of 500 kHz. 
In total the dataset consists of 22 species from five families, 449 individual bats and 1816 individual echolocation call recordings. The distribution of these call recordings across species is summarised in Table \ref{tab::dataset}.

Collen's \cite{collen2012evolution} Bat super-tree provided the phylogenetic tree of the recorded bat species, \(\mathbf{T}\).

@@

We are interested in modelling the evolution of \(\mathcal{E}^{\mathbf{t}}(\cdot)\), the function valued trait representing \(x_{lm}^{\mathbf{t}}(\cdot)\), with a phylogenetic GP. However, only 22 species of bat are represented in \(\mathbf{T}\). The relatively small size of this dataset presents challenges for the estimation of the kernel hyperparameters in (\ref{eqn:oukernel}). A short simulation study was performed to investigate the the accuracy of estimated hyperparameters for a phylogenetic GP over \(\mathbf{T}\).

We are not limited to a single observation at any given \(\mathbf{t}\). By repeatedly sampling at each observed \(\mathbf{t}\), larger samples can be obtained, improving the quality of the estimators \(\hat{\theta}\). With this in mind, 1000 independent, univariate phylogenetic GPs were simulated for each of \(n = \{1,2,4,8\}\) according to the kernel (\ref{eqn:oukernel}) with \(\theta = [1,50,1]^{\mathsf{T}}\), where \(n\) is the number of samples generated at each leaf node. The likelihood of each of these samples (\ref{eqn:t2mle}) is then maximised to give a type II maximum likelihood estimator \(\hat{\theta}\) and the results summarised in Table \ref{tab::ind_simulation}. This simulation study indicates that at least \(n = 4\) observations are needed at each leaf node to provide stable estimators \(\hat{\theta}\).

@@

Given the modelling assumptions made in Section \ref{sec:funrep} an unbiased estimator for \(\mathcal{E}^{\mathbf{t}}(\cdot)\) is the sample mean given by 
\begin{equation}
\hat{\mathcal{E}}^{\mathbf{t}}~(\cdot) = \frac{1}{l_{\mathbf{t}}} \sum_{l = 1}^{l_{\mathbf{t}}} \frac{1}{m_l} \sum_{m = 1}^{m_l} \hat{\mathcal{E}}_{lm}^{\mathbf{t}}(\cdot)
\label{eqn:est}
\end{equation}
where \(m_l\) is the total number of recordings for the \(l^{th}\) individual and \(l_{\mathbf{t}}\) is the number of individuals recorded from the species at \(\mathbf{t} \in \mathbf{T}\). However, Table \ref{tab::ind_simulation} indicates that 22 samples is not enough to obtain a stable \(\hat{\theta}\) by type II maximum likelihood estimation. We implement a resampling procedure to leverage multiple estimates for each \(\mathcal{E}^{\mathbf{t}}(\cdot)\) from the dataset. This will produce a stable estimator, \(\hat{\theta}\).

@@

Our analysis set \(n_r = 4\) and included 4 samples of \(\hat{\mathcal{E}}_r^{\mathbf{t}}(\mathbf{f})\) in each \(\hat{\mathcal{E}}_r\) for \(r = 1,\dots,1000\). This reflected the structure of the dataset, for which the minimum number of individuals per species was 4, and the results of the simulations study which showed that 4 observations per species provided reasonably stable estimates for \(\theta\). Note also that \(\mathbf{f} = [9,10, \dots, 212]^{\mathsf{T}}\), which reflects the spectrum of frequencies over which bats emit echolocation calls. \(\hat{\phi}_F\) was obtained by identifying the first six principal components, which accounted for approximately 85\% of the variation, in each  \(\hat{\mathcal{E}}_r\). By averaging over each sample, a single set of six stable, approximately orthogonal, basis functions were identified. These basis functions were then passed through Blaschke \& Wiscott's \cite{blaschke2004cubica} algorithm to produce a set of six independent basis functions for \(\mathcal{E}^\mathbf{t}(\cdot)\). Thus \(\hat{X}_r\), the matrix of mixing coefficients described by (\ref{eqn:inv}), the columns of which are modelled a phylogenetic OU processes, is obtained for each \(\hat{\mathcal{E}}_r\). \(\hat{\theta}_r\) is then the type II maximum likelihood estimator of (\ref{eqn:t2mle}) given \(\hat{\mathcal{E}}_r\). Table \ref{tab::bat_results} presents the results of the hyperparameter estimation procedure.

@@

To perform ancestral trait reconstruction for \(\mathcal{E}^{\mathbf{t}^*}(\cdot)\) the species level traits are estimated by (\ref{eqn:est}) and and the model hyperparameters are set to be the mean values of \(\theta_r\) reported in Table \ref{tab::bat_results}.

@@

\section{Conclusions and Further Work}
\label{sec:conc}
This preliminary analysis has developed a model for the evolution of echolocation in bats and identified a phylogenetic signal which allows the construction of a posterior predictive distribution for ancestral traits. The log energy spectral density has been identified as a trait representative of the echolocation call in bats. This trait, representing the energy intensity of the call across the frequency spectrum, is modelled as a series of independent components, combinations of energy intensities across the spectrum, each of which evolves according to a phylogenetic Ornstein-Uhlenbeck process. Estimating the hyperparameters governing these Ornstein-Uhlenbeck processes from observed traits provides an insight into the evolution of these traits. Each of the hyperparameters has an intuitive interpretation where \(\frac{\sigma_p}{\sigma_p + \sigma_n}\) indicates the proportion of variation in the sample accounted for by the phylogenetic distance between species, while \(\ell\) provides a measure of how quickly correlation along the phylogeny decays. We are working towards understanding what the results of this analysis could mean with respect to the evolution of echolocation in bats.

One particular limitation of the model is the representation of the echolocation call by a log energy spectral density. Echolocation calls have complex spectral and temporal structures, much of which is lost in the log energy spectral density representation. An alternative time-frequency representation, which preserves more of this structure, is the spectrogram. Modelling the evolution of bat echolocation calls with spectrograms, and implementing this model for a larger dataset of bat echolocation calls, is to be the subject of future research.

The interested reader can access the datasets and code used to produce these results through the R package 'sdsBAT' which is still under development and can be found at https://github.com/jpmeagher/sdsBAT.
\bibliographystyle{ws-rv-van}
\bibliography{../BatBiblio}
